---
title: "Project title"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

**Research Question:** How does race, age, location, presence of mental illness, and other factors impact the prevalence of police shootings in the United States?

**Introduction/Data**

**Methodology**

```{r}
#| label: load-data
#| message: false
library(tidyverse)
library(tidymodels)
police_shootings <- read_csv("data/police_shootings.csv")

```

```{r}
#| label: age-prevalence

police_shootings |>
  filter(Person.Age>0) |>
  ggplot(aes(x=Person.Age, color = Person.Race)) +
  geom_bar() +
  labs(Title = "Relationship Between Victim Age and Prevalence of Police 
       Shooting", x= "Victim Age", y = "Number of Reported Shootings", 
       legend = "Race")

```

```{r}
#| label: shootings-over-time
police_shootings |>
  ggplot(aes(x=Incident.Date.Year, fill = Person.Race)) +
  geom_bar() + 
  labs(title = "P")
  
```

```{r}
#| label: mental illness-race
police_shootings |>
  ggplot(aes(x= Person.Race , fill=Incident.Location.State)) +
  geom_bar()
```

```{r}
library(usmap)
library(ggplot2)

plot_usmap(regions = "states", color = "red")
```

\*\* Results \*\*

-- What are you finding out??

We are investigating and determining which of these factors (race, age, location, presence of mental illness) has the most impact on shootings in the United States.

Graph interpretation: The first graph shows the correlation between victim age and number of reported shootings. Victims who are between the ages of 25-37 years appear to experience the highest prevalence of police shootings. Indeed, the number of people between these ages are significantly more affected than people older and younger, indicating that age is an important factor. However, across the whole spectrum of ages, it is also clear that people of African American race are more disproportionately affected, suggesting that race is also a significant factor in determining the prevalence of police shootings.

Graph 2: In our second graph we investigated the relationship between race and shootings more specifically. Just as our last graph indicated, African Americans are significantly affected more by police shootings than any other race, having approximately 1000 shootings per year between 2015-2020. In addition, Asians were not too far behind with 750 shootings per year.

-- Make sure that this is connected to your research question. It is easy to get excited and explore your data set. Make sure that all results you generate are in-line with the research question you have asked.

\
